Presheaves, Representables and\\the Yoneda Lemma

### 2017-07-05 sheets/talk-10/sheet-10.tex

\begin{exercise}
  Let $\cat A$ be a small category. Prove that the representable functors \emph{generate} the presheaf category $[\cat A^\op,\Set]$ in the folowing sense: Given any objects $F,G\in[\cat A^\op,\Set]$ and natural transformations $\alpha,\beta\from F\to G$, if for every representable functor $H_A$ and natural transformation $\gamma\from H_A\to F$, one has $\alpha\of\gamma=\beta\of\gamma$, then $\alpha=\gamma$. Thus, the arrows in $[\cat A^\op,\Set]$ are determined by their effect on generalized elements based at representables.
\end{exercise}

@@

%   A functor $F\from \cat A\to\cat B$ \emph{preserves limits} if it preserves limits of shape $\mathbf I$ for all small categories $\mathbf I$.
% \end{definition}
\begin{exercise}
  Fill in some of the missing details in the proof that for any small category $\cat A$ the presheaf category $[A^\op,\Set]$ is cartesian closed:
  \begin{enumerate}
  \item Show that for any $A\in\cat A$, $H_A$ maps all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$. What is the dual statement?
    \hint{It suffices to show that this holds for all coproducts and coequalizers.}
  \item Let $F\from \cat A\to\Set$ be a presheaf. Show that the functor $-\times F\from [\cat A^\op,\Set]\to[\cat A^\op,\Set]$ preserves colomits. That is: For any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$,
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F.
    \end{align*}
    You may assume that $[\cat A^\op,\Set]$ has all limits. (In particular limits, which are functors $\cat A^\op\to\Set$, are taken ``pointwise'': $$\left(\lim_{\to I\in\mathbf I} D(I)\right)(A) \iso \lim_{\to I\in\mathbf I} \left(D(I)(A)\right)$$ for all shapes and diagrams.)
    \hint{Use the fact that $[\cat A^\op,\Set]$ has all limits to deduce that there is an arrow $\lim_{\to \mathbf I} (D\times F)\to (\lim_{\to \mathbf I} D)\times F$ and use the Yoneda principle to show that this is an isomorphism.}
  \end{enumerate}
\end{exercise}

>>>>

\begin{exercise}
  Let $\cat A$ be a small category. Prove that the representable functors \emph{generate} the presheaf category $[\cat A^\op,\Set]$ in the folowing sense: Given any objects $F,G\in[\cat A^\op,\Set]$ and natural transformations $\alpha,\beta\from F\to G$, if for every representable functor $H_A$ and natural transformation $\gamma\from H_A\to F$, one has $\alpha\of\gamma=\beta\of\gamma$, then $\alpha=\beta$. Thus, the arrows in $[\cat A^\op,\Set]$ are determined by their effect on generalized elements based at representables.
\end{exercise}

@@

%   A functor $F\from \cat A\to\cat B$ \emph{preserves limits} if it preserves limits of shape $\mathbf I$ for all small categories $\mathbf I$.
% \end{definition}
\begin{exercise}
  Fill in some of the missing details in the proof that for any small category $\cat A$ the presheaf category $[A^\op,\Set]$ is cartesian closed:
  \begin{enumerate}
  \item Show that for any $A\in\cat A$, $H_A$ maps all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$. What is the dual statement?
    \hint{It suffices to show that this holds for all coproducts and coequalizers.}
  \item Let $F\from \cat A\to\Set$ be a presheaf. Show that the functor $-\times F\from [\cat A^\op,\Set]\to[\cat A^\op,\Set]$ preserves colimits. That is: For any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$,
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F.
    \end{align*}
    You may assume that $[\cat A^\op,\Set]$ has all limits. (In particular limits, which are functors $\cat A^\op\to\Set$, are taken ``pointwise'': $$\left(\lim_{\to I\in\mathbf I} D(I)\right)(A) \iso \lim_{\to I\in\mathbf I} \left(D(I)(A)\right)$$ for all shapes and diagrams.)
    \hint{Use the fact that $[\cat A^\op,\Set]$ has all limits to deduce that there is an arrow $\lim_{\to \mathbf I} (D\times F)\to (\lim_{\to \mathbf I} D)\times F$ and use the Yoneda principle to show that this is an isomorphism.}
  \end{enumerate}
\end{exercise}

### 2017-07-05 slides/slides-10/notes-10.tex

    We still have to check naturality, i.e. that for any $B'\xrightarrow{g}B$ the following commutes:
    \begin{align*}
    \xymatrix{
    H_A(B)=\cat A(B,A) \ar[r]^{H_A(g)=-\of g} \ar[d]^{\widetilde a_B} & H_A(B')=\cat A(B',A) \ar[d]_{\widetilde a_B'} \\
    F(B) \ar[r]^{F(g)} & F(B').
                         } 
  \end{align*}
  Since this square is in $\Set$, it suffices to check $(\widetilde a_B'\of H_A(g))(f)=(F(g)\of\widetilde a_B)(f)$ for every $f\in H_A(B)=\cat A(B,A)$. Indeed, this holds:
  \begin{align*}
    (\widetilde a_B'\of H_A(g))(f)&=(F(f\of g))(a)\\
    &=(F(f)\of F(g))(a)&F \text{ functor}\\
    &=F(f)(F(g)(a))\\
    &=(F(g)\of\widetilde a_B)(f).
  \end{align*}
\item Let $a\in F(A)$. Clearly, we get $\widehat{\widetilde a}=a$:
  \begin{align*}
    \widehat{\widetilde a}=\widetilde a_A(1_A)=F(1_A)(a)=1_{F(A)}(a)=a.
  \end{align*}
\item[(d)-(f)] The remaining proof obligations are quite straight-forward to show and are thus left as an exercise.
  \end{enumerate}
\end{proof}

@@

For the proof we need the following lemma, which we are also going to use in the proof that presheaf categories are cartesian closed:
\begin{lemma}
  \label{lem:HApres}
  For any locally small category $\cat A$ and $A\in\cat A$, $H_A$ maps all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$.
\end{lemma}
Now the proof of proposition \ref{prop:dis} goes through smoothly:
\begin{proof}[Proof of proposition \ref{prop:dis}]
  We have 
  \begin{align*}
    \cat A(A\times (B+C),X)&\iso\cat A((B+C)\times A,X)\\
                           &\iso\cat A(B+C,X^A)&\cat A\text{ cartesian closed}\\
                           &\iso\cat A(B,X^A)\times \cat A(C,X^A) &\text{by lemma \ref{lem:HApres}}\\
                           &\iso\cat A(A\times B,X)\times \cat A(A\times C,X) &\cat A\text{ cartesian closed}\\
                           &\iso\cat A((A\times B)+(A\times C),X).&\text{by the dual of lemma \ref{lem:HApres}}
  \end{align*}

@@

  Next, we have to define a diagram $F(\cat A)\to[\cat A^\op,\Set]$ such that $F$ is the corresponding limit. We simply take $H_-\of\pi$ where $\pi\from F(\cat A)\to\cat A$ is the projection on $\cat A$ defined by 
  \begin{itemize}
  \item $\pi(a,A)=A$ on objects and
  \item $\pi(h)=h$ on arrows in $\cat A$ .
  \end{itemize}

>>>>

    We still have to check naturality, i.e. that for any $B'\xrightarrow{g}B$ the following commutes:
    \begin{align*}
    \xymatrix{
    H_A(B)=\cat A(B,A) \ar[r]^{H_A(g)=-\of g} \ar[d]^{\widetilde a_B} & H_A(B')=\cat A(B',A) \ar[d]_{\widetilde a_B'} \\
    F(B) \ar[r]^{F(g)} & F(B').
                         } 
  \end{align*}
  Since this square is in $\Set$, it suffices to check $(\widetilde a_B'\of H_A(g))(f)=(F(g)\of\widetilde a_B)(f)$ for every $f\in H_A(B)=\cat A(B,A)$. Indeed, this holds:
  \begin{align*}
    (\widetilde a_B'\of H_A(g))(f)&=(F(f\of g))(a)\\
    &=(F(g)\of F(f))(a)&F \text{ contravariant functor}\\
    &=F(g)(F(f)(a))\\
    &=(F(g)\of\widetilde a_B)(f).
  \end{align*}
\item Let $a\in F(A)$. Clearly, we get $\widehat{\widetilde a}=a$:
  \begin{align*}
    \widehat{\widetilde a}=\widetilde a_A(1_A)=F(1_A)(a)=1_{F(A)}(a)=a.
  \end{align*}
\item[(d)-(f)] The remaining proof obligations are quite straight-forward to show and are thus left as an exercise.
  \end{enumerate}
\end{proof}

@@

For the proof we need the following lemma, which we are also going to use in the proof that presheaf categories are cartesian closed:
\begin{lemma}
  \label{lem:HApres}
  For any locally small category $\cat A$ and $A\in\cat A$, $H_A$ maps all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$.
\end{lemma}
Now the proof of proposition \ref{prop:dis} goes through smoothly:
\begin{proof}[Proof of proposition \ref{prop:dis}]
  We have 
  \begin{align*}
    \cat A(A\times (B+C),X)&\iso\cat A((B+C)\times A,X)\\
                           &\iso\cat A(B+C,X^A)&\cat A\text{ cartesian closed}\\
                           &\iso\cat A(B,X^A)\times \cat A(C,X^A) &\text{by lemma \ref{lem:HApres}}\\
                           &\iso\cat A(A\times B,X)\times \cat A(A\times C,X) &\cat A\text{ cartesian closed}\\
                           &\iso\cat A((A\times B)+(A\times C),X).&\text{by lemma \ref{lem:HApres}}
  \end{align*}

@@

  Next, we have to define a diagram $F(\cat A)\to[\cat A^\op,\Set]$ such that $F$ is the corresponding limit. We simply take $H_-\of\pi$ where $\pi\from F(\cat A)\to\cat A$ is the projection on $\cat A$ defined by 
  \begin{itemize}
  \item $\pi(a,A)=A$ on objects and
  \item $\pi(f)=f$ on arrows in $\cat A$ .
  \end{itemize}
